Applications of Technology and Large Data in Statistics Education and Statistical Graphics

### 2015-05-31 DefenseSlides/DefenseSlidesCommitteeHour.Rnw
Another round of minor slide edits found during practicing

Parameters must be specified such that: $\beta_0 \le x_{(0)} \le x_{(n)} \le \beta_J$
where $x_{(1)}$ and $x_{(n)}$ are the minimum and maximum data values

@@

Random Rectangular Binning
  \begin{itemize}
    \item Define set of bin centers: $\{x_j^\ast \text{ }|\text{ } x_j^\ast > x_{j-1}^\ast \}$ for $j \in \{1,\dots,J\}$
    \item Assign points at random to surrounding bin centers
    \item Probabilities inversely proportional do distance from bin centers
  \end{itemize}

@@

Slight issue with spatial loss for random binning
    \begin{itemize}
        \item Many possible random assignments result in same reduced binned data
        \item Find assignment with minimum total spatial loss
        \item Call this \textit{net} spatial loss
    \end{itemize}

@@

@@

Standard Frequency Binning
  \begin{itemize}
    \item Create k frequency groups using equal frequency bin widths
  \end{itemize}  
\pause

@@

\begin{frame}
    \frametitle{Frequency Binning Specification}

3 Main Choices
  \begin{itemize}
    \item Number of frequency bins (k)
    \item Transform counts (yes/no)
    \item Binning algorithm (standard/quantile)
  \end{itemize}

@@

<<SkewCountFreqBinning, echo=FALSE, warning=FALSE, fig.width=11, fig.height=4, out.width='.99\\linewidth', fig.pos='h',fig.align='center',fig.show='hold'>>=
binout1 <- RectBin2d(datlist[[2]]$x,datlist[[2]]$y,0,0,5,5,type="standard")
freqgrps = 5
binoutFreqGroups <- freqBin(binout1, binType="standard", ncolor=freqgrps, logCount=FALSE)
binoutQuantFreqGroups <- freqBin(binout1, binType="quantile", ncolor=freqgrps, logCount=FALSE)
binoutLogFreqGroups <- freqBin(binout1, binType="standard", ncolor=freqgrps, logCount=TRUE)
binoutQuantLogFreqGroups <- freqBin(binout1, binType="quantile", ncolor=freqgrps, logCount=TRUE)
cols <- seq_gradient_pal(low="#56B1F7", high="#132B43", space="Lab")((0:freqgrps)/freqgrps)
# Plot with Quantile Binned Frequencies
labels = paste(unique(binoutQuantFreqGroups[[1]]$freqlabel), "\n", c("1st Pentile","2nd Pentile","3rd Pentile","4th Pentile","5th Pentile"), sep="")
p3 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutQuantFreqGroups[[1]],
            main="Quantile Freq Binning") +
    xlab("") + ylab("")+
  scale_fill_manual("Frequencies", values=cols, 
                      guide = guide_legend(label.position="bottom", label.hjust=0.5,  
                                           title.position = "top"),
                      labels=labels) + 
  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
                     legend.key.width = unit(1, "cm"),
                     panel.grid= element_blank(),
                     plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
                     legend.text=element_text(size=12) ) 
# Plot with standard binned Log Frequencies
labels = unique(binoutLogFreqGroups[[1]]$freqlabel)
p4 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutLogFreqGroups[[1]],
            main="Standard Log Freq Binning") +
    xlab("") + ylab("")+
  scale_fill_manual("Log Frequencies", values=cols, 
                      guide = guide_legend(label.position="bottom", label.hjust=0.5,  
                                           title.position = "top"),
                      breaks=unique(binoutLogFreqGroups[[1]]$freqgroup),
                      labels=labels) + 
  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
                     legend.key.width = unit(1, "cm"),
                     panel.grid= element_blank(),
                     plot.margin = unit(c(0.5,.1,.1,.1),"cm") ,
                     legend.text=element_text(size=12)) 
grid.arrange(p4,p3, nrow=1)
@

@@

\begin{frame}
    \frametitle{Practical Takeaways for Binned Scatterplots}

Spatial Binning
  \begin{itemize}
      \item Standard binning algorithm superior to random binning in spatial loss and computation speed
      \item Smaller bins superior for precisely displaying spatial information     
      \item Make bin dimensions integer multiples of data resolution
      \item Quality default for origin offset based on data resolution
  \end{itemize}  
\pause

Frequency Binning
  \begin{itemize}
    \item May be used to make frequency loss explicit
    \item Use 4 to 7 frequency bins
    \item Quantile frequency binning provides alternative display for skewed counts
  \end{itemize}

@@

\begin{frame}
    \frametitle{Shiny Database Sampler}    
  Goals
    \begin{itemize}
      \item Connect intro students with big data
      \item Bypass technical hurdles of big data
    \end{itemize} \pause
  Technical Components
    \begin{itemize}
      \item \textit{Data storage}: MySQL
      \item \textit{Computation}: \texttt{R}, \texttt{shiny}, \texttt{RMySQL}
      \item \textit{User interface}: accessed through web browser
    \end{itemize} \pause
  Interface Design
    \begin{itemize}
      \item Quality characteristics of software design
      \item Cognitive load theory
      \item Human centered design
      \item Pedagogical value and curricular integration
    \end{itemize}
\end{frame}

@@

\begin{frame}
    \frametitle{Applications}    
  Lab Assignment
    \begin{itemize}
      \item Groups assignment
      \item Questions on hypothetical sampling goals
      \item Use Shiny Database Samplers to gather and explore samples
    \end{itemize}
  Course Projects
    \begin{itemize}
      \item Group project
      \item Research question with focus on bivariate associations
      \item Shiny Database Sampler use by some for mock surveys
    \end{itemize}
\end{frame}

>>>>

Parameters must be specified such that: $\beta_0 \le x_{(1)} \le x_{(n)} \le \beta_J$
where $x_{(1)}$ and $x_{(n)}$ are the minimum and maximum data values

@@

Random Rectangular Binning
  \begin{itemize}
    \item Define set of bin centers: $\{x_j^\ast \text{ }|\text{ } x_j^\ast > x_{j-1}^\ast \}$ for $j \in \{1,\dots,J\}$
    \item Assign points at random to surrounding bin centers
    \item Probabilities inversely proportional to distance from bin centers
  \end{itemize}

@@

Slight issue with spatial loss for random binning
    \begin{itemize}
        \item Many possible random assignments result in same reduced binned data
        \item Find equivalent assignment with minimum total spatial loss
        \item Call this \textit{net} spatial loss
    \end{itemize}

@@

\begin{frame}
    \frametitle{Practical Takeaways for Spatial Binning}

% Spatial Binning
  \begin{itemize}
      \item Standard binning algorithm superior to random binning in spatial loss and computation speed
      \item Smaller bins superior for precisely displaying spatial information     
      \item Make bin dimensions integer multiples of data resolution
      \item Quality default for origin offset based on data resolution
  \end{itemize}

@@

Standard Frequency Binning
  \begin{itemize}
    \item Create k frequency groups using equal frequency bin widths
    \item Same uncertainty for each shade
  \end{itemize}  
\pause

@@

% %-------------------------------------------------
% 
% \begin{frame}
%     \frametitle{Frequency Binning Specification}
%   
% 3 Main Choices
%   \begin{itemize}
%     \item Number of frequency bins (k)
%     \item Transform counts (yes/no)
%     \item Binning algorithm (standard/quantile)
%   \end{itemize}  
% 
% \end{frame}

@@

<<SkewCountFreqBinning, echo=FALSE, warning=FALSE, fig.width=11, fig.height=4, out.width='.99\\linewidth', fig.pos='h',fig.align='center',fig.show='hold'>>=
binout1 <- RectBin2d(datlist[[2]]$x,datlist[[2]]$y,0,0,5,5,type="standard")
freqgrps = 5
binoutFreqGroups <- freqBin(binout1, binType="standard", ncolor=freqgrps, logCount=FALSE)
binoutQuantFreqGroups <- freqBin(binout1, binType="quantile", ncolor=freqgrps, logCount=FALSE)
binoutLogFreqGroups <- freqBin(binout1, binType="standard", ncolor=freqgrps, logCount=TRUE)
binoutQuantLogFreqGroups <- freqBin(binout1, binType="quantile", ncolor=freqgrps, logCount=TRUE)
cols <- seq_gradient_pal(low="#56B1F7", high="#132B43", space="Lab")((0:freqgrps)/freqgrps)
# Plot with Quantile Binned Frequencies
#labels = paste(unique(binoutQuantFreqGroups[[1]]$freqlabel), "\n", c("1st Pentile","2nd Pentile","3rd Pentile","4th Pentile","5th Pentile"), sep="")
labels = paste(c("[1,3]","(3,17]","(17,107]","(107,510]","(510,3150]"), "\n", c("1st Pentile","2nd Pentile","3rd Pentile","4th Pentile","5th Pentile"), sep="")
p3 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutQuantFreqGroups[[1]],
            main="Quantile Freq Binning") +
    xlab("") + ylab("")+
  scale_fill_manual("Frequencies", values=cols, 
                      guide = guide_legend(label.position="bottom", label.hjust=0.5,  
                                           title.position = "top"),
                      labels=labels) + 
  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
                     legend.key.width = unit(1, "cm"),
                     panel.grid= element_blank(),
                     plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
                     legend.text=element_text(size=12) ) 
# Plot with standard binned Log Frequencies
labels = unique(binoutLogFreqGroups[[1]]$freqlabel)
p4 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutLogFreqGroups[[1]],
            main="Standard Log Freq Binning") +
    xlab("") + ylab("")+
  scale_fill_manual("Log Frequencies", values=cols, 
                      guide = guide_legend(label.position="bottom", label.hjust=0.5,  
                                           title.position = "top"),
                      breaks=unique(binoutLogFreqGroups[[1]]$freqgroup),
                      labels=labels) + 
  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
                     legend.key.width = unit(1, "cm"),
                     panel.grid= element_blank(),
                     plot.margin = unit(c(0.5,.1,.1,.1),"cm") ,
                     legend.text=element_text(size=12)) 
grid.arrange(p4,p3, nrow=1)
@

@@

% %-------------------------------------------------
% 
% \begin{frame}
%     \frametitle{Practical Takeaways for Binned Scatterplots}
%   
% Frequency Binning
%   \begin{itemize}
%     \item May be used to make frequency loss explicit
%     \item Use 4 to 7 frequency bins
%     \item Quantile frequency binning provides alternative display for skewed counts
%   \end{itemize}  
% 
% \end{frame}

@@

\begin{frame}
    \frametitle{Shiny Database Sampler}    
  Goals
    \begin{itemize}
      \item Connect intro students with big data
      \item Bypass technical hurdles of big data
    \end{itemize} \pause
  Technical Components
    \begin{itemize}
      \item \textit{Data storage}: SQL
      \item \textit{User interface}: accessed through web browser
      \item \textit{Computation}: \texttt{R}, \texttt{shiny}, \texttt{RMySQL}
    \end{itemize} \pause
  Interface Design
    \begin{itemize}
      \item Quality characteristics of software design
      \item Cognitive load theory
      \item Human centered design
      \item Pedagogical value and curricular integration
    \end{itemize}
\end{frame}

@@

\begin{frame}
    \frametitle{Applications}

  Used within course activities for \textit{Stat 104: Intro to Statistics}

  Lab Assignment
    \begin{itemize}
      \item Groups assignment
      \item Questions on hypothetical sampling goals
      \item Use Shiny Database Samplers to gather and explore samples
    \end{itemize}
  Course Projects
    \begin{itemize}
      \item Group project
      \item Research question with focus on bivariate associations
      \item Shiny Database Sampler use by some for mock surveys
    \end{itemize}
\end{frame}
